% -*- mode: ConTeXt; encoding: utf-8; -*-
%
% This is free and unencumbered software released into the public domain.
%
% Anyone is free to copy, modify, publish, use, compile, sell, or
% distribute this software, either in source code form or as a compiled
% binary, for any purpose, commercial or non-commercial, and by any
% means.
%
% In jurisdictions that recognize copyright laws, the author or authors
% of this software dedicate any and all copyright interest in the
% software to the public domain. We make this dedication for the benefit
% of the public at large and to the detriment of our heirs and
% successors. We intend this dedication to be an overt act of
% relinquishment in perpetuity of all present and future rights to this
% software under copyright law.
%
% THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
% EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
% MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
% IN NO EVENT SHALL THE AUTHORS BE LIABLE FOR ANY CLAIM, DAMAGES OR
% OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE,
% ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR
% OTHER DEALINGS IN THE SOFTWARE.

\enableregime[utf]
\frenchspacing
\mainlanguage[en]

\setupbodyfont[schoolbook, 11pt]
\setupbodyfontenvironment[default][em=italic]

\setuppapersize[letter][letter]

\setuplayout
  [location=singlesided,
    width=fit,
    backspace=1.25in]

\setuppagenumbering
  [alternative=singlesided,
    location={footer, middle},
    way=bytext]

\setuphead
  [title]
  [align=middle]
\setuphead
  [subject]
  [align=middle,
    commandbefore={---\hskip0.6em},
    commandafter={\hskip0.5em---}]

\starttext

\title{Cookies for Alice and Bob}

\subject{The Problem}

\startitemize[n]
\item You have one box of Chocolate Cookies and one box of Vanilla
  Cookies.
\item You will deliver one of the boxes to Alice and one of the boxes
  to Bob. Neither has a preference of flavor.
\item You flip a coin to decide who gets which box.\footnote{This step
  is equivalent to a light source generating a pair of perpendicularly
  polarized photons.}
\item Alice has a favorite angle,~$\alpha$. Bob also has a favorite
  angle,~$\beta$. You write down the numbers $\cos^2\alpha$,
  $\sin^2\alpha$, $\cos^2\beta$, and~$\sin^2\beta$. Also, you have a
  pair of ten-sided dice. You can roll these dice to get a number
  between~$0.00$ and~$0.99$, inclusive.
\item Suppose Alice is getting the box of Chocolate Cookies. You roll
  the dice. If the number you roll is less than
  $\cos^2\alpha$, then you write~$+1$ on the box. Otherwise
  you write~$-1$.\footnote{This is equivalent to a photon being
    detected in one of the channels of a polarizing beam splitter.}
\item Suppose, on the other hand, that Alice is getting the box of
  Vanilla Cookies. You roll the dice.  If the number you roll is less
  than $\sin^2\alpha$, then you write~$+1$ on the
  box. Otherwise you write~$-1$.
\item Suppose Bob is getting the box of Chocolate Cookies. You roll
  the dice. If the number you roll is less than
  $\cos^2\beta$, then you write~$+1$ on the box. Otherwise
  you write~$-1$.
\item Suppose, on the other hand, that Bob is getting the box of
  Vanilla Cookies. You roll the dice.  If the number you roll is less
  than $\sin^2\beta$, then you write~$+1$ on the
  box. Otherwise you write~$-1$.
\item Calculate the correlation coefficient of the numbers written on
  the cookie boxes, as a function of~$\alpha-\beta$.\footnote{The
    answer from quantum mechanics is~$-\!\cos\,2(\alpha-\beta)$, and
    physicists have assured us that only by traversal of states in a
    Hilbert space can one derive this answer. Of course, this claim
    defies the principles of mathematics and is preposterous.}
\stopitemize

\subject{The Solution}

\startitemize[n]
\item First let us solve the problem
  for~$\alpha_0\in(-\infty,+\infty)$ and~$\beta_0=0$. Then
  \startformula
    \startmathalignment [n=2, align={right,left}]
      \NC\cos^2\beta_0\NC= 1 \NR
      \NC\sin^2\beta_0\NC= 0 \NR
    \stopmathalignment
  \stopformula
  and so the~$\beta_0=0$ setting does nothing except tell us to mark
  Chocolate Cookies with a~$+1$ and Vanilla Cookies with a~$-1$.

\item Thanks to our choice of markings, the correlation coefficient
  equals the covariance (that is, there is no need to normalize), and
  is
  \startformula
    \rho=\rho^{++}+\rho^{+-}+\rho^{-+}+\rho^{--}
  \stopformula
  where
  \startformula
    \startmathalignment [n=2, align={right,left}]
      % The roll for Bob always succeeds for Chocolate, always fails for
      % Vanilla.
      \NC\rho^{++}\NC=\frac{1}{2}(+1)(+1)\sin^2\alpha_0\NR % Vanilla/Choc
      \NC\rho^{+-}\NC=\frac{1}{2}(+1)(-1)\cos^2\alpha_0\NR % Choc/Vanilla
      \NC\rho^{-+}\NC=\frac{1}{2}(-1)(+1)\cos^2\alpha_0\NR % Vanilla/Choc
      \NC\rho^{--}\NC=\frac{1}{2}(-1)(-1)\sin^2\alpha_0\NR % Choc/Vanilla
    \stopmathalignment
  \stopformula

\item Thus (by a double-angle identity found in reference books)
  \startformula
    \rho=-(\cos^2\alpha_0-\sin^2\alpha_0)=-\cos\,2\alpha_0=-\cos\,2(\alpha_0-\beta_0)
  \stopformula
  and~$\rho$ is the solution for the special case~$\alpha_0$
  and~$\beta_0$.

\item Now set~$\alpha_0=\alpha-\beta$. That
  makes~$\alpha=\alpha_0+\beta$
  and~$\beta=\beta_0+\beta$. Add~$\beta-\beta=0$ to~$\alpha_0-\beta_0$
  in the last expression for~$\rho$ above. This gives
\startformula
  \rho=-\cos\,2(\alpha-\beta)
\stopformula
as the general solution.

\item Interestingly, although the particular ‘scattering’ of~$+1$
  and~$-1$ marks (were a person to do the operation many times)
  depends on the specific values of~$\alpha$ and~$\beta$, the
  correlation coefficient depends only on the
  difference~$\alpha-\beta$. That is, there exists a relationship
  between~$+1$ and~$-1$ marks that is invariant under in-unison
  rotation of the angles.\footnote{Of course, the same rotational
    invariance was always true of Bell test experiments, but must not
    have been widely noticed, for otherwise Clauser would not have
    received a Nobel Prize.}~$\square$
\stopitemize

\stoptext
